Exploring probability distributions for bivariate temporal granularities

### 2020-01-30 main.Rmd
cricket section making concise - removing the pattern across season

The application is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized in other applications. The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season can consist of several matches. The ball by ball data for IPL season 2008 to 2016 is fetched from [Kaggle](https://www.kaggle.com/josephgpinto/ipl-data-analysis/data). The `cricket` data set in the `gravitas` package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016), each of which has 6 balls in an over and 2 innings in a match.
s

@@

Although there is no conventional time granularity in cricket, we can still represent the data set `cricket` through a `tsibble`, with each over representing an ordering from past to future, can form the index of the tsibble. The hierarchy table would look like the following:

```{r hierarchy2}
library(gravitas)
library(tibble)
hierarchy_model <- tibble::tibble(
  G = c("over", "inning", "match", "season"),
  C = c("over-of-inning", "inning-of-match", "match-of-season",1),
  P = c(20, 2, "aperiodic", 1))
knitr::kable(hierarchy_model, format = "markdown")

@@

First, we look at the distribution of runs (measured variable) across over-of-inning (circular granularity) and match-of-season (aperidic cyclic granularity) in \autoref{fig:seas-over-inning}. The distribution of runs per over has not significantly changed from 2008 to 2016. There is no clear pattern/trend that runs per over is increasing or decreasing across seasons. Hence, we work with subsets of seasons to answer some of the questions:

@@

```{r runs_per-over}
library(tsibble)

cricket_all <-  read_csv("data-raw/deliveries_all.csv")
matches_all <- read_csv("data-raw/matches_all.csv")

cricket_season <- cricket_all %>% left_join(matches_all, by = c("match_id" ="id"))

cricket_per_over <- cricket_season %>% 
  group_by(season, match_id, batting_team, bowling_team,  inning, over) %>%
  summarise(runs_per_over = sum(total_runs), run_rate = sum(total_runs)/length(total_runs))

cricket_tsibble_all <- cricket_per_over %>% 
  ungroup() %>% 
  mutate(data_index = row_number()) %>% 
  as_tsibble(index = data_index)
```

```{r seas-over-inning, fig.cap = "Quantile plot of runs per over across overs of different seasons. There is no pattern on increase or decrease of runs across overs for seasons."}
  cricket_tsibble_all %>% 
  prob_plot("over", 
            "season",
            hierarchy_model,
            response = "runs_per_over",
            plot_type = "quantile",
            quantile_prob = c(0.25, 0.5, 0.75), 
            symmetric = FALSE)  +
     theme( # remove the vertical grid lines
            panel.grid.major.x = element_blank() ,
            # explicitly set the horizontal lines (or they will disappear too)
            panel.grid.major.y = element_blank()) + scale_x_discrete(breaks = seq(2008, 2016, 3))
```

>>>>

The method is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized to other applications. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. An idea of cyclic granularities can be conceived with this hierarchy in mind. Examples may include ball of the over, over of the innings or ball of the innings. Although most of these cyclic granularities are circular in theory by design of the hierarchy, in reality these may be aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when a inning finishes before the over finishes. Thus, the cyclic granularity ball-of-the-over will be circular in most cases and aperiodic in others. However, irrespective of the type of cyclic granularities, it can be interesting to visualize the periodicity of  scores/performance under similar data structure as temporal datasets.

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The ball by ball data for IPL season 2008 to 2016 is fetched from [Kaggle](https://www.kaggle.com/josephgpinto/ipl-data-analysis/data). The `cricket` data set in the `gravitas` package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each inning has 5 overs and each match has 2 innings.

@@

<!-- The hierarchy table would look like the following: -->

<!-- # ```{r hierarchy2} -->
<!-- # library(gravitas) -->
<!-- # library(tibble) -->
<!-- # hierarchy_model <- tibble::tibble( -->
<!-- #   G = c("over", "inning", "match", "season"), -->
<!-- #   C = c("over-of-inning", "inning-of-match", "match-of-season",1), -->
<!-- #   P = c(20, 2, "aperiodic", 1)) -->
<!-- # knitr::kable(hierarchy_model, format = "markdown") -->
<!-- #  -->
<!-- # ``` -->

@@

<!-- Making it short -->

<!-- First, we look at the distribution of runs (measured variable) across over-of-inning (circular granularity) and match-of-season (aperidic cyclic granularity) in \autoref{fig:seas-over-inning}. The distribution of runs per over has not significantly changed from 2008 to 2016. There is no clear pattern/trend that runs per over is increasing or decreasing across seasons. Hence, we work with subsets of seasons to answer some of the questions: -->

@@

<!-- # ```{r runs_per-over} -->
<!-- # library(tsibble) -->
<!-- #  -->
<!-- # cricket_all <-  read_csv("data-raw/deliveries_all.csv") -->
<!-- # matches_all <- read_csv("data-raw/matches_all.csv") -->
<!-- #  -->
<!-- # cricket_season <- cricket_all %>% left_join(matches_all, by = c("match_id" ="id")) -->
<!-- #                                          -->
<!-- # cricket_per_over <- cricket_season %>%  -->
<!-- #   group_by(season, match_id, batting_team, bowling_team,  inning, over) %>% -->
<!-- #   summarise(runs_per_over = sum(total_runs), run_rate = sum(total_runs)/length(total_runs))  -->
<!-- #  -->
<!-- # cricket_tsibble_all <- cricket_per_over %>%  -->
<!-- #   ungroup() %>%  -->
<!-- #   mutate(data_index = row_number()) %>%  -->
<!-- #   as_tsibble(index = data_index) -->
<!-- # ``` -->
<!-- #  -->
<!-- #  -->
<!-- # ```{r seas-over-inning, fig.cap = "Quantile plot of runs per over across overs of different seasons. There is no pattern on increase or decrease of runs across overs for seasons."} -->
<!-- #   cricket_tsibble_all %>%  -->
<!-- #   prob_plot("over",  -->
<!-- #             "season", -->
<!-- #             hierarchy_model, -->
<!-- #             response = "runs_per_over", -->
<!-- #             plot_type = "quantile", -->
<!-- #             quantile_prob = c(0.25, 0.5, 0.75),  -->
<!-- #             symmetric = FALSE)  + -->
<!-- #      theme( # remove the vertical grid lines -->
<!-- #             panel.grid.major.x = element_blank() , -->
<!-- #             # explicitly set the horizontal lines (or they will disappear too) -->
<!-- #             panel.grid.major.y = element_blank()) + scale_x_discrete(breaks = seq(2008, 2016, 3)) -->
<!-- # ``` -->
